\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in R

### 2014-01-13 papers/jss/article.Rnw
pass 3
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/rprotobuf@766 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

The previous sections discussed functionality in the \pkg{RProtoBuf} package
for creating, manipulating, parsing and serializing Protocol Buffer messages.
In addition to these low-level methods, the package also has some high level
functionality for automatically converting R data structures into protocol
buffers and vice versa. The \texttt{serialize\_pb} and \texttt{unserialize\_pb}
functions convert arbitrary R objects into a universal Protocol Buffer structure:

@@

In order to accomplish this, \pkg{RProtoBuf} uses the same catch-all \texttt{proto}
schema that \pkg{RHIPE} uses for exchanging R data with Hadoop \citep{rhipe}. This 
schema, which we will refer to as \texttt{rexp.proto} is printed in appendix 
\ref{rex.proto}. Even though the \texttt{RHIPE} implementation is written in Java and
\texttt{RProtoBuf} is writting in R and \texttt{C++}, the Protocol Buffer messages
are naturally compatible between the two systems because they use the same schema. 
This shows the power of using a schema based cross-platform format such as Protocol 
Buffers: interoperability is archieved without tight coordination or collaboration.

@@

There are \Sexpr{n} standard data sets included in the base-r \pkg{datasets}
package. These datasets include data frames, matrices, time series, tables lists,
and some more exotic data classes. The \texttt{can\_serialize\_pb} method can be 
used to determine which of those can fully be converted to the \texttt{rexp.proto}
Protocol Buffer representation:

<<echo=TRUE>>=
datasets$valid.proto <- sapply(datasets$name, function(x) can_serialize_pb(get(x)))
datasets <- subset(datasets, valid.proto==TRUE)
m <- nrow(datasets)
@

\Sexpr{m} data sets can be converted to Protocol Buffers
(\Sexpr{format(100*m/n,digits=1)}\%) without loss of information. Upon closer
inspection, all other datasets are objects of class \texttt{nfnGroupedData}.
This class represents a special type of data frame that has some additional 
attributes used by the \pkg{nlme} package, among which a \emph{formula} object.
Because formulas are R \emph{language} objects, they have little meaning to
other systems, and are not supported by the \texttt{rexp.proto} descriptor.
When \texttt{serialize\_pb} is used on objects of this class (or other objects
containing unsupported data types), it will serialize all other values and 
attributes of the object, but skip over the unsupported types with a warning.

@@

TODO comparison of protobuf serialization sizes/times for various vectors.
Compared to R's native serialization.  Discussion of the RHIPE approach of
serializing any/all R objects, vs more specific Protocol Buffers for specific
R objects.

@@

@@

As described earlier, the primary application of Protocol Buffers is data
interchange in the context of inter-system communications.  Network protocols
such as HTTP provide mechanisms for client-server communication, i.e. how to
initiate requests, authenticate, send messages, etc.  However, many network
protocols generally do not regulate the \emph{content} of messages: they
allow transfer of any media type, such as web pages, static files or
multimedia content.  When designing systems where various components require
exchange of specific data structures, we need something on top of the network
protocol that prescribes how these structures are to be represented in
messages (buffers) on the network. Protocol Buffers solve exactly this
problem by providing a cross-platform method for serializing arbitrary
structures into well defined messages, which can then be exchanged using any
protocol. The descriptors (\texttt{.proto} files) are used to formally define
the interface of a remote API or network application. Libraries to parse and
generate protobuf messages are available for many programming languages,
making it relatively straightforward to implement clients and servers.

@@

Because both HTTP and Protocol Buffers have libraries available for many 
languages, clients can be implemented in just a few lines of code. Below
is example code for both R and Python that retrieves a dataset from R with 
OpenCPU using a protobuf message. In R, we use the HTTP client from 
the \texttt{httr} package \citep{httr}.
% superfluous?
%, and the protobuf parser from the \texttt{RProtoBuf} package.
In this example we
download a dataset which is part of the base R distribution, so we can
verify that the object was transferred without loss of information.

@@

This code suggests a method for exchanging objects between R servers, however this can 
also be done without Protocol Buffers. The main advantage of using an inter-operable format 
is that we can actually access R objects from within another
programming language. For example, in a very similar fashion we can retrieve the same
dataset in a Python client. To parse messages in Python, we first compile the 
\texttt{rexp.proto} descriptor into a python module using the \texttt{protoc} compiler:

@@

OpenCPU also provides a lot of meta-functionality such as handling
of sessions, exceptions, security, and more. OpenCPU also makes it possible to store
output of a function call on the server, instead of directly retrieving it. Thereby 
objects can be shared with other users or used as arguments in a subsequent
function call. But in its essence, the HTTP API provides a simple way to perform remote 
R function calls over HTTPS. The same request can be performed in Python as follows:

@@

The MapReduce programming model \citep{dean2008mapreduce} has emerged
in the last decade as a popular framework for processing large data
sets in parallel on large compute clusters.  Protocol Buffers
provide a convenient mechanism to send structured data between tasks
in a MapReduce cluster.  In particular, the large data sets in fields
such as particle physics and information processing are frequently
stored in binned or histogram form in order to reduce the data storage
requirements for later data analysis \citep{scott2009multivariate}.

>>>>

The previous sections discussed functionality in the \pkg{RProtoBuf} package
for creating, manipulating, parsing and serializing Protocol Buffer messages.
In addition to these low-level methods, the package also has some high level
functionality for automatically converting R data structures into protocol
buffers and vice versa. The \texttt{serialize\_pb} and \texttt{unserialize\_pb}
functions serialize arbitrary R objects into a universal Protocol Buffer 
message:

@@

In order to accomplish this, \pkg{RProtoBuf} uses the same catch-all \texttt{proto}
schema used by \pkg{RHIPE} for exchanging R data with Hadoop \citep{rhipe}. This 
schema, which we will refer to as \texttt{rexp.proto} is printed in appendix 
\ref{rexp.proto}. The Protocol Buffer messages generated by \pkg{RProtoBuf} and
\pkg{RHIPE} are naturally compatible between the two systems because they use the 
same schema. This shows the power of using a schema based cross-platform format such
as Protocol Buffers: interoperability is archieved without effort or close coordination.

The \texttt{rexp.proto} schema supports all main R storage types holding \emph{data}.
These include \texttt{NULL}, \texttt{list} and vectors of type \texttt{logical}, 
\texttt{character}, \texttt{double}, \texttt{integer} and \texttt{complex}. In addition,
every type can contain a named set of attributes, as is the case in R. The \texttt{rexp.proto}
schema does not support some of the special R specific storage types, such as \texttt{function},
\texttt{language} or \texttt{environment}. Such objects have no native equivalent 
type in Protocol Buffers, and have little meaning outside the context of R.
When serializing R objects using \texttt{serialize\_pb}, values or attributes of
unsupported types are skipped with a warning. If the user really wishes to serialize these 
objects, they need to be converted into a supported type. For example, the  can use 
\texttt{deparse} to convert functions or language objects into strings, or \texttt{as.list}
for environments.

@@

There are \Sexpr{n} standard data sets included in the base-r \pkg{datasets}
package. These datasets include data frames, matrices, time series, tables lists,
and some more exotic data classes. The \texttt{can\_serialize\_pb} method is 
used to determine which of those can fully be converted to the \texttt{rexp.proto}
Protocol Buffer representation. This method simply checks if any of the values or
attributes in an object is of an unsupported type:

<<echo=TRUE>>=
m <- sum(sapply(datasets$name, function(x) can_serialize_pb(get(x))))
@

\Sexpr{m} data sets can be converted to Protocol Buffers
(\Sexpr{format(100*m/n,digits=1)}\%) without loss of information. Upon closer
inspection, all other datasets are objects of class \texttt{nfnGroupedData}.
This class represents a special type of data frame that has some additional 
attributes used by the \pkg{nlme} package, among which a \emph{formula} object.
Because formulas are R \emph{language} objects, they have little meaning to
other systems, and are not supported by the \texttt{rexp.proto} descriptor.
When \texttt{serialize\_pb} is used on objects of this class, it will serialize
the data frame and all attributes, except for the formula.

<<>>=
attr(CO2, "formula")
msg <- serialize_pb(CO2, NULL)
object <- unserialize_pb(msg)
identical(CO2, object)
identical(class(CO2), class(object))
identical(dim(CO2), dim(object))
attr(object, "formula")
@

@@

@@

%JO: is this section really relevant? Maybe just a citation will do instead?

@@

As described earlier, the primary application of Protocol Buffers is data
interchange in the context of inter-system communications.  Network protocols
such as HTTP provide mechanisms for client-server communication, i.e. how to
initiate requests, authenticate, send messages, etc.  However, network
protocols generally do not regulate the \emph{content} of messages: they
allow transfer of any media type, such as web pages, static files or
multimedia content.  When designing systems where various components require
exchange of specific data structures, we need something on top of the network
protocol that prescribes how these structures are to be represented in
messages (buffers) on the network. Protocol Buffers solve exactly this
problem by providing a cross-platform method for serializing arbitrary
structures into well defined messages, which can then be exchanged using any
protocol. The descriptors (\texttt{.proto} files) are used to formally define
the interface of a remote API or network application. Libraries to parse and
generate protobuf messages are available for many programming languages,
making it relatively straightforward to implement clients and servers.

@@

Because both HTTP and Protocol Buffers have libraries available for many 
languages, clients can be implemented in just a few lines of code. Below
is example code for both R and Python that retrieves a dataset from R with 
OpenCPU using a protobuf message. In R, we use the HTTP client from 
the \texttt{httr} package \citep{httr}. In this example we
download a dataset which is part of the base R distribution, so we can
verify that the object was transferred without loss of information.

@@

This code suggests a method for exchanging objects between R servers, however this might as 
well be done without Protocol Buffers. The main advantage of using an inter-operable format 
is that we can actually access R objects from within another
programming language. For example, in a very similar fashion we can retrieve the same
dataset in a Python client. To parse messages in Python, we first compile the 
\texttt{rexp.proto} descriptor into a python module using the \texttt{protoc} compiler:

@@

OpenCPU also provides a lot of meta-functionality such as handling
of sessions, exceptions, security, and more. OpenCPU also makes it possible to store
output of a function call on the server, instead of directly retrieving it. Thereby 
objects can be shared with other users or used as arguments in a subsequent
function call. But in its essence, the HTTP API provides a simple way to perform remote 
R function calls over HTTPS. The same request can be performed in Python as demonstrated
below. The code is a bit verbose because to show how the REXP message is created from 
scratch. In practice would probably write a function or small module construct a Protocol
Buffer message representing an R list from a Python dictionary object.

@@

Over the past years, the MapReduce programming model \citep{dean2008mapreduce}
has emerged as a poweful design pattern for processing large data
sets in parallel on large compute clusters.  Protocol Buffers
provide a convenient mechanism to send structured data between tasks
in a MapReduce cluster.  In particular, the large data sets in fields
such as particle physics and information processing are frequently
stored in binned or histogram form in order to reduce the data storage
requirements for later data analysis \citep{scott2009multivariate}.
